---
title: "Fossil Teeth"
output: pdf_document
---

For all fossil teeth; 232 observations (teeth collected). Teeth are only
distinguished by ID number *no morphologic sorting involved


```{r}

Fossildata <- read.delim("~/Shark-Morphometrics/Shark Data/Fossilsteeth.txt")

```

#PCA
```{r} 

#since we do not have any known data about the fossil teeth (jaw type, position, etc.)
#the only ordination approach which we can test is the PCA, since it is unconstrained.

#we would really like to look for trends with regards to jaw type and possibly 
#tooth position, based on the trends we previously observed across ALL taxa in the modern sharks

par(mfrow=c(1,2))
fossilcords <- Fossildata[ , 3:30]
row.names(fossilcords)
pca_fossil = princomp(fossilcords)  #pca of all coordainates
biplot(pca_fossil, cex=.5)

text(pca_fossil$scores, as.character(Fossildata$individual), cex=.7)

pca_fossil$loadings
summary(pca_fossil)

# overlap fossil teeth on extant PCA
allcoords = ExtantData[ , 7:34]
row.names(allcoords)
pca_shark = princomp(allcoords)  #pca of all coordainates
biplot(pca_shark, cex=.5, choices = c(1, 2))
pca_shark$loadings
summary(pca_shark)

fos_pre = predict(pca_shark, newdata=fossilcords)
points(fos_pre[ , 1], fos_pre[ , 2], col='blue', pch=19)


#this is showing that when the fossil teeth are plotted within the PCA of the extant, it encompases the variation but the 
#principle componenets one and 2 do not encompass the majority of the variation explained by the fossils. 

# pca with both extant and fossil teeth
library(vegan)
allteeth = rbind(fossilcords, allcoords)
pca_all = rda(allteeth ~ 1)
plot(pca_all, display='sites', pch='.')
all_pts = scores(pca_all, 1:2, 'sites')
fossil_pts = all_pts[1:nrow(fossilcords), ]
extant_pts = all_pts[(nrow(fossilcords)+1):nrow(all_pts), ]
points(fossil_pts, col='red')
points(extant_pts, col='blue')

summary(pca_all)
#when we include the PCs together, we can see a similar distribution of points, such that
# it may be possible that this is explained by jaw type and or tooth posiition (both?)


#what is interesting is that the direction of procrustes coordinates go in the same direction
#lower jaw is the top cluster, upper jaw is the bottom cluster
```

Just by looking at the PCA distribution of points, it looks like we have 4 
distinct clusters. Clear separation (possible jaw type) and then small groups within 
the larger one, side of jaw?

#Pam algorithm and Kmeans
```{r}

fossilcords
fossilmatrix <- as.matrix(fossilcords)

#Pam method (more robust than kmeans)

library("cluster", lib.loc="~/R/x86_64-pc-linux-gnu-library/3.2")
pam(Fossildata[,3:30],4, cluster.only=TRUE)
plot(pam(x=Fossildata[ ,3:30], k=4), main="", col.p="black")

#would be cool to look at this partitional clustering plot with the points their individual # to have an 
#idea of where they may cluster too (to go back and observe.)

```
The basic pam algorithm is fully described in chapter 2 of Kaufman and Rousseeuw(1990). Compared to the k-means approach in kmeans, the function pam has the following features: (a) it also accepts a dissimilarity matrix; (b) it is more robust because it minimizes a sum of dissimilarities instead of a sum of squared euclidean distances; (c) it provides a novel graphical display, the silhouette plot (see plot.partition) (d) it allows to select the number of clusters using mean(silhouette(pr)[, "sil_width"]) on the result pr <- pam(..), or directly its component pr$silinfo$avg.width, see also pam.object.( page 123 R book)


#elbow method for kmeans
```{r}

totvariance <- sum(diag(var(fossilmatrix[, 1:28])))*232
SSratio2 <- numeric(10)
 for (i in 1:10)
 {mod <- kmeans(fossilmatrix[, 1:28],i)
   SSratio2[i]<-(totvariance-sum(mod$withinss))/totvariance}
   

plot(1:10, SSratio2, xlab="number of clusters", ylab="% of explained variance")
points(1:10, c(0, SSratio2[2:10]), pch=3)

#4 or 5 clusters looks like it appears to describe the most variation
#4 is consistent with the PAM algorithm

```

This is confirming how many clusters it is best to use (which is 4)



#distance matrix of fossil teeth to extant ones



























