---
title: "Shark Teeth Full MkDn"
output: pdf_document
---

#Read in the Shark data;  separating out the 3 species and jaw type for each sp.

```{r}
library(vegan)
library(dummies)

sharkdat <- read.delim("~/Shark-Teeth-Morphometrics/sharkdat.txt")
View(sharkdat)

```

#Subset species

```{r} 

#Subset species and jaw type.
#The three species below are in the genus
#Carcharhinus or requiem sharks. 

altimus<-sharkdat[sharkdat$species=="Ca",]
brevipinna<-sharkdat[sharkdat$species=="Cb",]
falciformis<-sharkdat[sharkdat$species=="Cf",]

#subset of upper and lower jaws (includes all species)

upper<-sharkdat[sharkdat$jaw.type=="U",]
lower<-sharkdat[sharkdat$jaw.type=="L",]

upper_coord<- upper[, 6:33]
lower_coord <- lower[, 6:33]

#as.factor jaw and tooth from the entire dataset

jaw <- as.factor(sharkdat$jaw.type) #2 levels, upper and lower
tooth <- as.factor(sharkdat$tooth.position) #15 levels, tooth position


```

#Loop to edit data frame into matrix

```{r}

#fucntion for transforming data frame into a matrix

get_xydat = function(ct_mat) {
    if (class(ct_mat) != 'matrix') {
        ct_mat = as.matrix(ct_mat)
        print('Converting input to a matrix')
    }
    nmarkers =  ncol(ct_mat) /2 
    xy_mat = matrix(NA, nrow= nrow(ct_mat) * nmarkers, ncol=4)
    colnames(xy_mat) = c('indiv', 'landmark', 'x','y')
    startrow = 1
    for(i in 1:nrow(ct_mat)) { 
        x = ct_mat[i, seq(1, ncol(ct_mat) - 1, 2)]
        y = ct_mat[i, seq(2, ncol(ct_mat), 2)]
        endrow = i * nmarkers
        xy_mat[startrow:endrow , ] = cbind(rep(i, nmarkers), 1:nmarkers, x,y)
        startrow = endrow + 1
    }
    xy_mat = as.data.frame(xy_mat)
    return(xy_mat)
}

```

#All Species Coordinate Data

```{r}

#all coordinates for entire shark data set

coords = sharkdat[ , 6:33]
all_shark_coords <- get_xydat(coords)

head(all_shark_coords) 

#this has put all teeth as it's own individual. We have 5 animals consisting of 3 species, and each animal has 30 teeth (15 upper and 15 lower). Each tooth has 14 landmarks found on it, represented by coordinates x and y. 

```

Species and Jaw: I want to find the mean x and y coordinates for each landmark position (14) for all upper jaw teeth, and lower jaw teeth for each species. 
averaging individual specimen and tooth position for both jaw types 
for each species.)

#just coordinates of upper and lower jaws for each species; take out metadata

```{r}

#ALTIMUS

#subset jaws: upper and lower for altimus

upper_alt <- altimus[altimus$jaw.type=="U",]
lower_alt <-  altimus[altimus$jaw.type=="L",]

upper_alt_coord <- upper_alt[ ,6:33 ]
lower_alt_coord <- lower_alt[ ,6:33 ]


altimus_coords <- altimus[ ,6:33]
xydat_alti_all <- get_xydat(altimus_coords) #xy data for all altimus teeth

upper_alt_coord <- upper_alt[ ,6:33 ]
xydat_alti_upper <- get_xydat(upper_alt_coord)

lower_alt_coord <- lower_alt[ ,6:33 ]
xydat_alti_lower <- get_xydat(lower_alt_coord)


#BREVIPINNA

#subset upper and lower jaws for brevipinna

upper_brevi <- brevipinna[brevipinna$jaw.type=="U",]
lower_brevi <-  brevipinna[brevipinna$jaw.type=="L",]

brevipinna_coords <- brevipinna[ , 6:33]
xydat_brevi_all <-  get_xydat(brevipinna_coords) #xy data for all brevi teeth

upper_brevi_coord <- upper_brevi[ ,6:33]
xydat_brevi_upper <-  get_xydat(upper_brevi_coord) #xy data upper brevi teeth

lower_brevi_coord <- lower_brevi[ ,6:33]
xydat_brevi_lower <- get_xydat(lower_brevi_coord) #xy data lower brevi teeth


#FALCIFORMIS

#subset upper and lower jaws for falciformis

upper_falci <- falciformis[falciformis$jaw.type=="U",]
lower_falci <- falciformis[falciformis$jaw.type=="L",]

falciformis_coords <- falciformis[ ,6:33]
xydat_falci_all = get_xydat(falciformis_coords) #xy data for all falci

upper_falci_coord <- upper_falci[ ,6:33]
xydat_falci_upper <- get_xydat(upper_falci_coord) #xy data for just upper jaw of #falci

lower_falci_coord <- lower_falci[ ,6:33]
xydat_falci_lower <- get_xydat(lower_falci_coord) #xy data for lower jaw of falci

```




#Mean Upper and Lower Jaw Coordinates for Altimus. 

```{r}
#find the mean of x and y for each landmark position
#mean coordinates for altimus upper jaw

x_avg_alt_upper = tapply(xydat_alti_upper$x, xydat_alti_upper$landmark, mean)
y_avg_alt_upper = tapply(xydat_alti_upper$y, xydat_alti_upper$landmark, mean)


#mean coordinates for lower jaw of altimus

x_avg_alt_lower <- tapply(xydat_alti_lower$x, xydat_alti_lower$landmark, mean)
y_avg_alt_lower <- tapply(xydat_alti_lower$y, xydat_alti_lower$landmark, mean)

#This gives you the mean coordinate for each landmark found on both the #upper and lower jaw teeth. Allows us to plot the overall shape (next step)
#mean shape, and analyze the distribution of landmarks between tooth 
#positions for upper and lower jaw teeth of altimus. This can be done
#for all jaws for all species, but this is just for visual sake. 

```


##average shape of Upper and lower jaw of Car. Altimus

```{r}
#Upper Jaw

plot(x_avg_alt_upper, y_avg_alt_upper, pch=19, col='red', xlim=range(xydat_alti_upper$x), ylim=range(xydat_alti_upper$y)) #plot of mean shape. mean of each landmark
range(xydat_alti_upper$x)

points(xydat_alti_upper$x, xydat_alti_upper$y) #all landmarks ploted
plot(xydat_alti_upper$x, xydat_alti_upper$y)
points(x_avg_alt_upper, y_avg_alt_upper, pch=19, col='red') #mean plotted in red

#plot the mean shape of upper jaw teeth of altimus in Red, and the black 
#points represent the distribution of coordinates around each landmark.
#these are due to positional variation (position of tooth in the jaw. 
#indicated in data set by tooth.position in sharkdat)


#do the same for the Lower Jaw

plot(x_avg_alt_lower, y_avg_alt_lower)
plot(x_avg_alt_lower, y_avg_alt_lower, pch=19, col='red')

points(xydat_alti_lower$x, xydat_alti_lower$y)
plot(x_avg_alt_lower, y_avg_alt_lower, pch=19, col='red', xlim=range(xydat_alti_lower$x), ylim=range(xydat_alti_lower$y))
range(xydat_alti_lower$x)

points(xydat_alti_lower$x, xydat_alti_lower$y)
plot(xydat_alti_lower$x, xydat_alti_lower$y)
points(x_avg_alt_lower, y_avg_alt_lower, pch=19, col='red')

#when we do the same thing for the lower jaw of altimus, we can see that
#(visually) that the mean shape changes due to the jaw type *lower* and 
#the distribution of coordinates around each landmark changes. 

#we could do this same visual technique to analyze shape change for each species,
#but the focus is to look at shape change between the species. Not positional 
#change.

```

##RDA; Redundancy Analysis

since I am working with multivariate data (14 response landmarks each with their own x/y coordinates), would like to extract those gradients of variation in tooth shape (landmarks) expalinable by the independent variables (tooth position, individual, species, and jaw type)
The RDA serves the purpose of a MANOVA for my dataset .

#What is the effect of species, jaw type, and tooth position on overall variation?

```{r}

#run an RDA to look at variance partitioning

all_rda <- rda(coords ~ sharkdat$species + as.factor(sharkdat$jaw.type) +
                 as.factor(sharkdat$tooth.position))

#after looking at several RDA's (looking at interactions and inclusion of #individuals) , this RDA outputs the most variance explained by the #independent variables. 

anova(all_rda, permutations = 1000)
anova(all_rda, by="margin")

#output shows that jaw type accounts for a significant amount of variation, followed 
#by species and then tooth position. 


RsquareAdj(all_rda)

#variance explained by the RDA is about 58%. This is fairly good. 

plot(all_rda, type='n', scaling=1)
orditorp(all_rda, display='sp', cex=0.5, scaling=1, col='blue')
text(all_rda, display='bp', col='red')

#VarPart for Variance Partitioning

sp_dum = dummy(sharkdat$species)
jaw_dum = dummy(sharkdat$jaw.type)
pos_dum = dummy(sharkdat$tooth.position)

varpart(coords, sp_dum, jaw_dum, pos_dum)
showvarparts(3)


#this is saying that the jaw explains about 30% of the variance, tooth position
#explains 18% of the variance, and species only 7%. Since these 3 species #are 
#closely related in the genus Carcharhinus, they show little #differentiation. 
#The upper and lower jaws show the most amount of variance, since they have
#different functions, as well as the tooth position. 


```

The RDA allows us to see what predictor variables account for variation
but for a visual, let's analyze the distribution of raw data in a PCA; see if there is an emphasis in variation among the individual samples by gradients of the total sample variance. Visually Analyze structure of distribution of sample variance 

------------------------------------------------------------------------
#Ordination Methods
##PCA

#can teeth be quantitatively distinguished from one another?

```{r}

#Run PCA on all of the shark data; 3 species

coords <- sharkdat[ , 6:33]
row.names(coords) <- sharkdat$Id
pca_shark = princomp(coords)  #pca of all coordainates
biplot(pca_shark, cex=.5)
pca_shark$loadings
summary(pca_shark)
#PCA 1 and 2 explain about 67% of the overall variance which also does a very
#good job. however, PCA does not base this off of a priori groups. 



plot(pca_shark)

#Here we can see a significant amount of structure occuring in our raw #data. We can see that there are 2 obvious clusters within the overall
#raw data, the upper jaw teeth of all species cluster separately from 
#the lower jaw teeth of all the species. 

#this leads us to see if there is clustering within the lower and upper
#jaw teeth separated from one another. #there are too many data points in 
#the full data set to see if species cluster together separately in the 
#upper and lower jaws. 


#look at pca of upper and lower jaw separate to see if species separate

upper_coord
row.names(upper_coord) <- upper$Id
pca_upper <- princomp(upper_coord)
biplot(pca_upper, cex = .5)

lower_coord
row.names(lower_coord) <- lower$Id
pca_lower <- princomp(lower_coord)
biplot(pca_lower, cex= .5)


#when separating out the full pca to upper and lower pca's, we can see
#that there is some clustering within the upper and lower jaws based on #species, specifically in the upper jaw. We can see that altimus and #falciformis are clustering together in some areas of the same tooth #position. So there is overalp going on , showing similar tooth shape #between these species , but there is also some clustering of species #occurring. Ca appears to occupy the upper poriton of the graph, while Cf #occupies the left and center protion of the graph, and Cb occupies the #lower porition of the graph (all of these for the most part). However, #there is significant overlap occuring in the centroid of the graph, so 
#let's see if there is another analysis that will allow us to look at 
#species separation/overlap for the upper and lower jaws. 

```

Here we can see the PCA has a clear distinction between the upper and lower jaws
, I want to try and see if a CVA will be able to distinguish between species too. 


##CVA; canonical variate analysis

canonical variates analysis - A method of multivariate analysis in which the variation among groups is expressed relative to the pooled within-group covariance matrix. Canonical variates analysis finds linear transformations of the data which maximize the among group variation relative to the pooled within-group variation. The canonical variates then may be displayed as an ordination to show the group centroids and scatter within groups.The canonical variates are uncorrelated, however the vectors of coefficients are not orthogonal as in Principal Component Analysis.A critical assumption is that the within-group variance-covariance structure is similar, otherwise the pooling of the data over groups is not very sensible.

##Cva using Package Morpho; CVA of Upper and lower jaws as a predictor

(this is what the RDA states as explaining the most variance)

```{r}

install.packages("Morpho")
library(Morpho)

coords <- sharkdat[ , 6:33]
factor <- as.factor(paste(sharkdat$species, sharkdat$jaw.type, sep="_"))

#RUN the CVA

cva_1 <- CVA(coords, groups=factor, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = FALSE, p.adjust.method = "none")

cva_1$Var
cva_2$Dist

# get the typicality probabiliites and classifications - tagging specimens with a low probability < 0.01
#as an outlie (unassigned) 

typprobs <- typprobClass(cva_1$CVscores,groups=factor)
print(typprobs)

```

#Plotting the CV

```{r}


#visualize CV scores by their groups 

plot(cva_1$CVscores, col=factor, pch=as.numeric(factor), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_1$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_1$Var[2,1],1),"%")))

text(cva_1$CVscores, as.character(factor), col=as.numeric(factor), cex=.7)

#add elipses
     for(ii in 1:length(levels(factor))){
        dataEllipse(cva_1$CVscores[factor==levels(factor)[ii],1],
          cva_1$CVscores[factor==levels(factor)[ii],2], 
          add=TRUE,levels=.80, col=c(1:7)[ii])
  }

#This does a much better job than the PC in distinguishing groups of species
#for the upper and lower jaws. It also explains 20% more variance than the first
#two pC's do. 


library(lattice)

#histogram showing the distribution of CA's among the species
#and jaw types. explains 66% of overall variaion

histogram(~cva_1$CVscores[,1]|factor,
  layout=c(1,length(levels(factor))),
  xlab=paste("1st canonical axis", paste(round(cva_1$Var[1,2],1),"%")))

#histogram showing same thing for CA 2; 20% of overall variation

histogram(~cva_1$CVscores[,2]|factor, 
      layout=c(1,length(levels(factor))),
      xlab=paste("2nd canonical axis", paste(round(cva_1$Var[2,2],1),"%")))


#plot Mahalahobis; this will show the relationships between my groups

dendroS=hclust(cva_1$Dist$GroupdistMaha)
dendroS$labels=levels(factor)
par(mar=c(4,4.5,1,1))
dendroS=as.dendrogram(dendroS)
plot(dendroS, main='',sub='', xlab="Shark Species and Jaw",
     ylab='Mahalahobis distance')


# Variance explained by the canonical roots:

cva_1$Var
# or plot it:
barplot(cva_1$Var[,2])

#CV's do a very good job of explaining overall variation in the dataset.
#much better than the PCA. 

```

-------------------------------------------------------------------------------

##LDA of entire (all) matrix by species
#are some teeth better than others than distinguishing among species

```{r}

library(MASS)


upper_coord<- upper[, 6:33]
lower_coord <- lower[, 6:33]

jaw_type <-as.factor(upper[ ,4])


#make a matrix of all coords, then separate matrix to upper and lower.

matshark <- as.matrix(coords)


#overall lda

shark_lda <- lda(matshark, as.factor(sharkdat[ ,2]))
plot(shark_lda, col = 'blue')

#this does a pretty nice job of separating species out from one another
#But we can't tell what is the upper and lower jaw. Continue to do an 
#lda that distinguishes the upper and lower jaws from one another. 
#this is the opposite result of the pca; it does a better job at 
#distinguishing between species but not jaw type.

```

#analyze the lda with the upper and lower jaws separated
#LDA of Species, Upper and Lower Jaws Separated. 

```{r}

upper_coord<- upper[, 6:33]
lower_coord <- lower[, 6:33]

matshark_upper <- as.matrix(upper_coord)
matshark_lower <- as.matrix(lower_coord)

 
#Upper

upper_lda <- lda(matshark_upper, as.factor(upper[ ,2]))
plot(lda(matshark_upper, as.factor(upper[ ,2])))

#Here are the upper jaws, and the discriminant function is able to cluster #species
#distinct from one another from the upper jaw teeth. Upper jaw teeth shape #are a clasifier
#for identification to species
#we can see that the lda detected collinear variables, which is probably why it 
#couldn't distinguish between jaws and species. 


#Lower

lower_lda <- lda(matshark_lower, as.factor(lower[ ,2]))
plot(lda(matshark_lower, as.factor(lower[ ,2])))

#Here we can see that the altimus species have completely different lower #teeth
#but the lower teeth of falciformis and brevipinna have lots of overlap, 
#due to the fact they exhibit similar shape. So lower jaw teeth are not as #good
#for identifying down to the species level based on tooth shape because #they
#share more overlap. 


```

#Mahalanobis Distance Between Upper jaws species and lower jaws species

It takes into consideration the variance and correlation of the variables in measuring distances between points, i. e., differences in directions in which there is less variation within groups are given greater weight than are differences in directions in which there is more variation.

```{r}

uppermean <- upper_lda$mean
dist(predict(upper_lda, uppermean)$x)

#the mahaalanobis distance from each of the species shows they are 
#easily distinguished from one another (the larger the number, the greater 
#the greater the difference between the grous)


lowermean <- lower_lda$mean
dist(predict(lower_lda, lowermean )$x)

#here there is significant difference between Ca and Cb as expressed by the
#mahalanobis distance, along with Ca and Cf, however Cb and Cf are quite similar
#due to the small distance measurement. THis shows that there is significant overlap
#and similarities between the shape of lower jaw teeth of falciformis and #brevipinna. So there are not as good at predicting species by the shape of the
#lower jaw teeth. 

```

##Mahalanobis Distance Between Shark Species


```{r}

#now let's compute the distances between mean groups in space defined by the #dis.functions
#the compute the mahalanobis distance between groups (separating our species)

meangroup <- shark_lda$mean
dist(predict(shark_lda, meangroup)$x)

#if we compute the same for the overall LDA of the data set, we can see the same
#overlap occuring between Cf and Cb, and we can still see the clustering of Ca #separate from the other 2 species. The differences are skewed in here because
#this is combining both the shape data for the upper and lower jaws. The previous
#calculation (distance) separting out the jaw type is a better depiction 
#of analyzing the shape data. 

```


#Clustering our Data Set; Cluster Analysis

This emphasizes both differences and similarities among individual clusters
by clustering these entities based on inter-entity resemblance. 

#find optimal number of clusters for dataset

```{r}

#this function will output a log function graph showing the optimal amount
#of clusters that describe the most variation found in the dataset. 

p<-as.matrix(coords)

totv <- sum(diag(var(p[, 1:28])))*388
SSratio <- numeric(10)
 for (i in 2:10)
 {mod <- kmeans(p[, 1:28],i)
   SSratio[i]<-(totv-sum(mod$withinss))/totv}
   

plot(1:10, SSratio, xlab="number of clusters", ylab="% of explained variance")
points(1:10, c(0, SSratio[2:10]), pch=3)

#we can see here that the amount of clusters that will explain the most 
#amount of variance is about 6 (where it reaches the threshhold.)
```

#apply cluster function to dataset

will emphasize both differences and similarities among the samples by clustering
entities based on inter-group resemblance. Plot 6 clusters into the function.

```{r}
library(cluster)
library("clusterGeneration", lib.loc="~/R/win-library/3.1")
pam(p[,1:28], 6)$clustering

plot(pam(x=p[,1:28], k=6), main= "", col.p="black")

#While messy, we can see how there are 6 clusters that explain most of the 
#variance in the data set (refer to PCA loadings). These clusters show 
#those formed from the upper and lower jaws of each species. (3 species)
#Notice significant overlap which is most likely due to species similarity,
#but we can see some good separation most likely due to jaw type (which 
#does explain the most variation as seen in our RDA). So let's evaluate clustering
#due to just 2 clusters, upper and lower jaw type. So this analysis confirms
#our previous findings in the LDA analysis. Let's see if it does a better
#job when we separate just to 2 clusters, which we saw in the PCA. 


#Examine the same plot but examine Upper vs. Lower Teeth overlap
#Upper teeth are indicated with a number 2, lower teeth with a number 1. 
#Lower teeth in the first position are categorized in the upper jaw group
#because they have a indentical shape form to those in the upper jaw.
#clear separation is seen, overlap is due to similar teeth shapes of the first position teeth
#of lower jaw with upper. 

pam(x=p[,1:28], 2)$clustering
plot(pam(x=p[,1:28], k=2), main= "", col.p="black")


#confirms our previous findings that the jaws are the better descriptor 
#of variation than are species. 
#from here, using the upper jaws (since it is better to detect differences between species, let's analyze to see which landmark contributes to the most change
#between the species, so which might be the cause of shape change between these 3 #spcies. )


```


Since the upper jaw teeth are the most identifiable to species (more variable),
let's 
analyze where shape change is occuring in a single tooth relative to species.
In a previous  study, Naylor noted that the 5th position tooth in the upper 
jaw is better to distinguish among species. By looking at just one tooth in the 
upper jaw
, we are now limiting shape change now to due to species. I want to see if 
if there is visual shape change across these 3 species, and what landmarks
are associated with shape change shown by a large standard deviation. 

----------------------------------------------------------------------------
##Extract Tooth Position 5 for the Upper Jaw for 3 species
#what landmarks attribute to the most variation in shape among species (euclidean distance)
```{r}

upper<-sharkdat[sharkdat$jaw.type=="U",]
fifth <- upper[upper$tooth.position=="5",]

#extract just the 5th tooth for each species 
a5<-fifth[fifth$species=="Ca",]
b5<-fifth[fifth$species=="Cb",]
f5<-fifth[fifth$species=="Cf",]


#just coordinates
a5_coords <- a5[ ,6:33]
b5_coords <- b5[ ,6:33]
f5_coords <- f5[ ,6:33]

#mean of all 5th position
m5_coords <- fifth[ ,6:33]

#xy data matrix
xydat_a5 <- get_xydat(a5_coords)
xydat_b5 <- get_xydat(b5_coords)
xydat_f5 <- get_xydat(f5_coords)

xydat_m5 <- get_xydat(m5_coords)

#altimus
x_avg_a5 <- tapply(xydat_a5$x, xydat_a5$landmark, mean)
y_avg_a5 <-tapply(xydat_a5$y, xydat_a5$landmark, mean)

#brevipinna
x_avg_b5 <- tapply(xydat_b5$x, xydat_b5$landmark, mean)
y_avg_b5 <-tapply(xydat_b5$y, xydat_b5$landmark, mean)

#falciformis
x_avg_f5 <- tapply(xydat_f5$x, xydat_f5$landmark, mean)
y_avg_f5 <-tapply(xydat_f5$y, xydat_f5$landmark, mean)


#mean
x_avg_m5 <- tapply(xydat_m5$x, xydat_m5$landmark, mean)
y_avg_m5 <- tapply(xydat_m5$y, xydat_m5$landmark, mean)
```


#Plot the 5th position for each species

```{r}

#altimus
plot(x_avg_a5, y_avg_a5, pch=19, col='red') #mean plotted in red

#brevi
points(x_avg_b5, y_avg_b5, pch=19, col='blue')

#falci
points(x_avg_f5, y_avg_f5, pch=19, col='green')

#mean shape of all 
points(x_avg_m5, y_avg_m5, pch=19, col='grey')

```

Here we can see shape changes associated with a single tooth position (5)of the 
upper jaw across 3 species and their deviation away from the mean shape. THe 
mean was calculated as the average of all the 5th position teeth from the 
3 species put together. 

##Calculate the distance between those landmarks

The distance between two landmarks is the square root of the sum of the 
squared differences between each landmark. (Euclidean distance)
This allows us to calculate a single value distance from one landmark to the 
next in each species, so then we can use these to compare to the next species.
We are going to calculate the Euclidean distance because it is a coordinate free approach for comparing biological shapes using landmark data.

```{r}

#matrix of each species, upper jaw

alti <- cbind(x_avg_a5, y_avg_a5) #red
brevi <- cbind(x_avg_b5, y_avg_b5) #blue
falci <- cbind(x_avg_f5, y_avg_f5) #grenen
mean_5 <- cbind(x_avg_m5, y_avg_m5) #grey

#calculate the interlandmark distance between each landmark, for each species
#and for the mean
#(i.e: distance between landmark 1>2>3>4>5>6...14)

dist_alti <- vegdist(alti, method="euclidean")
dist_alti <- vegdist(brevi, method="euclidean")
dist_alti <- vegdist(falci, method="euclidean")

dist_mean <- vegdist(mean_5, method="euclidean")

```

compare to see how these distances differ from the mean (which landmarks attribute
to the greatest change per/ species.) 

##Euclidean distance of species from mean (upper jaw, 5th pos.)

I want to see what landmarks are attributing to the most change in tooth shape
for each species ( from mean reference shape, to target shape:species)

```{r}

#difference of altimus from mean
apply(mean_5 - alti,1,function(x){sqrt(x[1]^2+x[2]^2)})

#difference of brevipinna from mean
apply(mean_5 - brevi,1,function(x){sqrt(x[1]^2+x[2]^2)})

#difference of falciformis from mean
apply(mean_5 - falci,1,function(x){sqrt(x[1]^2+x[2]^2)})


adist = apply(mean_5 - alti,1,function(x){sqrt(x[1]^2+x[2]^2)})
bdist = apply(mean_5 - alti,1,function(x){sqrt(x[1]^2+x[2]^2)})
bdist = apply(mean_5 - brevi,1,function(x){sqrt(x[1]^2+x[2]^2)})
fdist = apply(mean_5 - falci,1,function(x){sqrt(x[1]^2+x[2]^2)})
cbind(adist,bdist,fdist)

apply(cbind(adist,bdist,fdist),1,sd)

#so we can see that the 4th landmark (the tip of the blade) contributes to
#the most change in shape with regards to species in the upper jaw.
#we can see a considerable amount of variation due to landmark 3, 10, and 13 as well. 

```