---
title: "Extant Jaws (Right)"
output: pdf_document
---

Read in the Procrustes Coordinates from ALL extant shark species right side 
of jaw

#Read in Data
```{r}
Extant.Jaw.Sets.R.data <- read.delim("~/Shark-Morphometrics/Shark Data/Extant Jaw Sets R data.txt")

ExtantData<-Extant.Jaw.Sets.R.data
View(ExtantData)

#this data is found on my desktop for all extant species/jaws . right side only

```

#Subset Species and Jaws
```{r}
acronotus<-   ExtantData[ExtantData$species=="Ca",]
altimus<-     ExtantData[ExtantData$species=="ca",]
brevipinna<-  ExtantData[ExtantData$species=="Cb",]
falciformis<- ExtantData[ExtantData$species=="Cf",]
longimanus<-  ExtantData[ExtantData$species=="Cg",]
leucas<-      ExtantData[ExtantData$species=="Cu",]
prionace<-    ExtantData[ExtantData$species=="Pg",]
lewini<-      ExtantData[ExtantData$species=="Sl",]
rhizo<-       ExtantData[ExtantData$species=="Rt",]
isodon<-      ExtantData[ExtantData$species=="Ci",]
limbatus<-    ExtantData[ExtantData$species=="Cl",]
obscurus<-    ExtantData[ExtantData$species=="Co",]
plumbeus<-    ExtantData[ExtantData$species=="Cp",]
taurus<-      ExtantData[ExtantData$species=="Ct",]
white<-       ExtantData[ExtantData$species=="Gw",]
oxy<-         ExtantData[ExtantData$species=="Io",]
tiburo<-      ExtantData[ExtantData$species=="St",]
paucus<-      ExtantData[ExtantData$species=="Ip",]

#subset of upper and lower jaws (includes all species)

upperjaw<-ExtantData[ExtantData$jaw.type=="U",]
lowerjaw<-ExtantData[ExtantData$jaw.type=="L",]

```


#Loop to edit data frame into matrix

```{r}

#fucntion for transforming data frame into a matrix

get_xydat = function(ct_mat) {
    if (class(ct_mat) != 'matrix') {
        ct_mat = as.matrix(ct_mat)
        print('Converting input to a matrix')
    }
    nmarkers =  ncol(ct_mat) /2 
    xy_mat = matrix(NA, nrow= nrow(ct_mat) * nmarkers, ncol=4)
    colnames(xy_mat) = c('indiv', 'landmark', 'x','y')
    startrow = 1
    for(i in 1:nrow(ct_mat)) { 
        x = ct_mat[i, seq(1, ncol(ct_mat) - 1, 2)]
        y = ct_mat[i, seq(2, ncol(ct_mat), 2)]
        endrow = i * nmarkers
        xy_mat[startrow:endrow , ] = cbind(rep(i, nmarkers), 1:nmarkers, x,y)
        startrow = endrow + 1
    }
    xy_mat = as.data.frame(xy_mat)
    return(xy_mat)
}

```

#All Species Coordinate Data

```{r}

#all coordinates for entire shark data set in xydat form

allcoords = ExtantData[ , 7:34]
all_coords<- get_xydat(allcoords)

head(all_coords) 

upper_coords <- upperjaw[, 7:34]
lower_coords <- lowerjaw[, 7:34]

```

#RDA
RDA can also be considered a constrained version of principal components analysis (PCA), wherein canonical axes - built from linear combinations of response variables - must also be linear combinations of the explanatory variables . is a method to extract and summarise the variation in a set of response variables that can be explained by a set of explanatory variables. More accurately, RDA is a direct gradient analysis technique which summarises linear relationships between components of response variables that are "redundant" with (i.e. "explained" by) a set of explanatory variables (very similar to a MANOVA)

```{r}
#What is the effect of species, jaw type, and tooth position on overall variation?


#run an RDA to look at variance partitioning

extant_rda <- rda(allcoords ~ ExtantData$species + as.factor(ExtantData$jaw.type) +
                 as.factor(ExtantData$tooth.position))


anova(extant_rda, permutations = 1000)
anova(extant_rda, by="margin")

#intuitively, know that jaw type has a significant amount of difference in tooth
#shape. which it does, and species and tooth position , express the same amounts of
#variation (just about). interesting

RsquareAdj(extant_rda)

#66% of overall variation explained. pretty nice amount


#VarPart for Variance Partitioning

spec_dum = dummy(ExtantData$species)
jawty_dum = dummy(ExtantData$jaw.type)
tooth_dum = dummy(ExtantData$tooth.position)

varpart(allcoords, spec_dum, jawty_dum, tooth_dum)

#again, jaw type has the most variation at 33% and species + jaw type have almost
#equal amounts of variation 

showvarparts(3)

#not super useful to look at

```

##PCA

```{r}

#Run PCA on all of the shark data

allcoords = ExtantData[ , 7:34]
row.names(allcoords)
pca_shark = princomp(allcoords)  #pca of all coordainates
biplot(pca_shark, cex=.5)
pca_shark$loadings
summary(pca_shark)
#PCA 1 and 2 explain about 70% of the overall variance which also does a very
#good job. however, PCA does not base this off of a priori groups. I wanted
#to be able to see if species could be differentiated from one another. 



plot(pca_shark)

#Here we can see a significant amount of structure occuring in our raw #data. We can see that there are 2 obvious clusters within the overall
#raw data, the upper jaw teeth of all species cluster separately from 
#the lower jaw teeth of all the species. The exact same figure that we get 
#from the morphoJ package. 

#this leads us to see if there is clustering within the lower and upper
#jaw teeth separated from one another. #there are too many data points in 
#the full data set to see if species cluster together separately in the 
#upper and lower jaws. 

```


#PCA by tooth position
doesn't give me the output I'm looking for. (did this with CVA in MorphoJ)
```{r}

acronotus <- ExtantData[ExtantData$species=="Ca",]
upper_acronotus <- acronotus[acronotus$jaw.type=="U",]
lower_acronotus <- acronotus[acronotus$jaw.type=="L",]

upper_acronotus_coord <- upper_acronotus[ ,7:34 ]
lower_acronotus_coord <- lower_acronotus[ ,7:34 ]

acronotus$tooth.position

testpca$loadings


```


#CVA
```{r}

##Cva using Package Morpho; CVA of Upper and lower jaws as a predictor

#(this is what the RDA states as explaining the most variance)

install.packages("Morpho")
library(Morpho)


allcoords = ExtantData[ , 7:34]

factor <- as.factor(paste(ExtantData$species, ExtantData$jaw.type, sep="_"))


#RUN the CVA

cva_1 <- CVA(allcoords, groups=factor, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = FALSE, p.adjust.method = "none")


cva_1$Var
cva_1$Dist

# get the typicality probabiliites and classifications - tagging specimens with a low probability < 0.01
#as an outlier (unassigned) 

typprobs <- typprobClass(cva_1$CVscores,groups=factor)
print(typprobs)

#the classification accurracy of the CV is about 86%. This is extremely high
#and shows that the separation among the teeth for each species is correct 
#86% of the time. 

#I need to see Dan about debugging the 0's in this to get a better estimate of the
#classification accuracy

```

#Plotting the CV

```{r}

install.packages("car")
require(car)

#visualize CV scores by their groups 

plot(cva_1$CVscores, col=factor, pch=as.numeric(factor), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_1$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_1$Var[2,1],1),"%")))

text(cva_1$CVscores, as.character(factor), col=as.numeric(factor), cex=.7)

#add elipses
     for(ii in 1:length(levels(factor))){
        dataEllipse(cva_1$CVscores[factor==levels(factor)[ii],1],
          cva_1$CVscores[factor==levels(factor)[ii],2], 
          add=TRUE,levels=.80, col=c(1:7)[ii])
  }

#This does a much better job than the PC in distinguishing groups of species
#for the upper and lower jaws. It also explains 20% more variance than the first
#two pC's do. However, it's pretty messy. I want to look at this for just upper and just for lower jaws. 


#plot Mahalahobis; this will show the relationships between my groups

sharkdendroS=hclust(cva_1$Dist$GroupdistMaha)
sharkdendroS$labels=levels(factor)
par(mar=c(4,4.5,1,1))
sharkdendroS=as.dendrogram(sharkdendroS)
plot(sharkdendroS, main='',sub='', xlab="Shark Species and Jaw",
     ylab='Mahalahobis distance')

#this is pretty cool. Look how the lamniforms cluster completely different from the
#carcharhinus sharks. Sand tiger is the outlier out of all of the species analyzed. 

cva_1$Dist

# Variance explained by the canonical roots:

cva_1$Var
# or plot it:
barplot(cva_1$Var[,2])

#CV's do a very good job of explaining overall variation in the dataset.
#much better than the PCA. 65% of cumulative % explained. not bad

```

#CVA by tooth position

To look at changes with respect to tooth
position

I did this in morphoJ with species, jaw type, and tooth position as classifiers.
The strongest one with the most signal is with species, jaw type, and tooth position
if we color the data points just with respect to jaw type, we can see clear separation
then if we color the data points with respect to species, we can see a lot of 
overlap, as expected but if we color the points with respect to tooth position
we see the progression of the reciprocal movement in the points with 
respect tooth position.

Variation among groups:
This CVA explains 70% of the overall variation!! Hey, that's pretty awesome.

*still trying to figure out the R script for this part.

```{r}
toothset <-ExtantData[ExtantData$species %in% c("Ip", "Io", "Ca", "ca", "Cb", "Cf", "Pg,", "Ci", "Cu","Sl", "Cl", "Cg", "Co", "Cp", "Rt", "St" ), ]

toothchords <- toothset[ ,7:34]
factor2 <- as.factor(paste(toothset$tooth.position, toothset$jaw.type, sep="_"))

cva_try <- CVA(toothset[ ,7:34], groups=factor2, weighting = TRUE, plot = TRUE,
              rounds = 0, cv = FALSE, p.adjust.method = "none")


```



#LDA Upper jaws by species (need to do this again for lower jaw) 

It takes into consideration the variance and correlation of the variables in measuring distances between points, i. e., differences in directions in which there is less variation within groups are given greater weight than are differences in directions in which there is more variation. Similar to PCA but does a better job of separating things.

```{r}

toothset <-ExtantData[ExtantData$species %in% c("Ip", "Io", "Ca", "ca", "Cb", "Cf", "Pg,", "Ci", "Cu","Sl", "Cl", "Cg", "Co", "Cp", "Rt", "St" ), ]

upperset<-toothset[toothset$jaw.type=="U",]
lowerset<-toothset[toothset$jaw.type=="L",]

chordsU <- upperset[,7:34]
chordsL <-lowerset[,7:34]

matrix_upper <- as.matrix(chordsU)
matrix_lower <- as.matrix(chordsL)

practice_lda <- lda(matrix_upper, as.factor(upperset[ ,2]))


plot(lda(matrix_upper, as.factor(upperset[ ,2])), lower.panel= NULL)

#LD1 and LD2 explain 60% of the variation. Just want to view this single plot (and possibly LD3-10%)



#now I want to look at plots individually, but I don't know how?

#this is basically an analysis to confirm the upper jaw teeth are better descriptors
#of species specific tooth shape , with regards to the Maha. distances, but has a better
#visual representation than the CVA 

uppermean <- practice_lda$mean
dist(predict(practice_lda, uppermean)$x)


```



Look at bivariate plots of the species that had little
separation (small differences). Doing something similar but also want to look at
the Lamnid lineage. (Similar to Naylor 1994)

#Evaluate distinct clusers; lamnid (unique group) and poor separation groups

```{r}

#lamnid lineage jaws separated

lamnid<- ExtantData[ExtantData$species %in% c("Ip", "Io", "Gw"), ]

lamnid_coords <- lamnid[ , 7:34]

lamnid_factor <- as.factor(paste(lamnid$species, lamnid$jaw.type, sep="_"))

cva_lamnid <- CVA(lamnid_coords, groups=lamnid_factor, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = TRUE, p.adjust.method = "none")

plot(cva_lamnid$CVscores, col=factor, pch=as.numeric(lamnid_factor), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_lamnid$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_lamnid$Var[2,1],1),"%")))

text(cva_lamnid$CVscores, as.character(lamnid_factor), col=as.numeric(lamnid_factor), cex=.7)

```

#poor separation in lower jaws of Cg, Cl, ca, Cp
```{r}


lowerjaw<-ExtantData[ExtantData$jaw.type=="L",]


Requ<-lowerjaw[lowerjaw$species %in% c("Cg", "Cl", "ca", "Cp"), ]
Requ_coords<- Requ[ , 7:34]

Requ_factor <- as.factor(paste(Requ$species, sep="_"))

cva_Requ <- CVA(Requ_coords, groups=Requ_factor, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = TRUE, p.adjust.method = "none")

plot(cva_Requ$CVscores, col=Requ_factor, pch=as.numeric(Requ_factor), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_Requ$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_Requ$Var[2,1],1),"%")))

text(cva_Requ$CVscores, as.character(Requ_factor), col=as.numeric(Requ_factor), cex=.7)

cva_Requ$Dist

#interestingly enough, even though this group of tooth shape shows the smallest separation,
#they are still distinct from one another, with the exception of "ca"

```

#poor separation in the upper jaws of Cu, Cg, and Co

```{r}

upperjaw<-ExtantData[ExtantData$jaw.type=="U",]


ReqU<-upperjaw[upperjaw$species %in% c("Cu", "Cg", "Co"), ]
ReqU_coords<- ReqU[ , 7:34]

ReqU_factor <- as.factor(paste(ReqU$species, sep="_"))

cva_ReqU <- CVA(ReqU_coords, groups=ReqU_factor, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = TRUE, p.adjust.method = "none")

plot(cva_ReqU$CVscores, col=ReqU_factor, pch=as.numeric(ReqU_factor), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_ReqU$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_ReqU$Var[2,1],1),"%")))

text(cva_ReqU$CVscores, as.character(ReqU_factor), col=as.numeric(ReqU_factor), cex=.7)

cva_ReqU$Dist


#again, pretty decent separation with little overlap. 

```


##LDA (all sharks)

```{r}

library(MASS)

#make a matrix of all coords, then separate matrix to upper and lower.

allcoords = ExtantData[ , 7:34]
matrixshark <- as.matrix(allcoords)

#jaw type lda

lda <- lda(matrixshark, as.factor(ExtantData[ ,4]))
plot(lda, col = 'blue')

#again, just shows nice separation with jaw type. Car. taurus is quite different.

```

##Mahalanobis Distance ^ LDA


```{r}

#now let's compute the distances between mean groups in space defined by the #dis.functions
#the compute the mahalanobis distance between groups (separating our species; this output
#give for jaw type averaged)

ldajaw <- lda(matrixshark, as.factor(ExtantData[ ,2]))
meangroup <- ldajaw$mean
dist(predict(ldajaw, meangroup)$x)


```


#Shape by tooth position (for distinct species)

refer to morphoJ project with PCA by tooth position
```{r}
#acronotus

acronotus<-   ExtantData[ExtantData$species=="Ca",]
upper_acronotus <- acronotus[acronotus$jaw.type=="U",]

shark2<- upper_acronotus[12:22,]

#upper jaw

plot(shark2$ProcCoord7~shark2$tooth.position, main="tooth position by shape upper", 
  	xlab="position", ylab="shape")

#lower jaw

lower_acronotus <- acronotus[acronotus$jaw.type=="L",]
shark3<- lower_acronotus[12:22,]

plot(shark3$ProcCoord7~shark3$tooth.position, main="tooth position by shape lower", 
  	xlab="position", ylab="shape")


#leucas

upper_leucas <- leucas[leucas$jaw.type=="U",]
bull1<-upper_leucas[12:22,]

lower_leucas <- leucas[leucas$jaw.type=="L",]
bull2<- lower_leucas[12:22,]

#upper jaw
plot(bull1$ProcCoord7~bull1$tooth.position, main="tooth position by shape upper", 
  	xlab="position", ylab="shape")

#lower jaw
plot(bull2$ProcCoord7~bull2$tooth.position, main="tooth position by shape lower", 
  	xlab="position", ylab="shape")
```


#all sharks , just the 4x landmark
```{r}
#this shows the shape trend for the upper jaw teeth and the lower jaw teeth
#by tooth position as you move from the anterior to the posterior part of the jaw
#for all sharks in the study. These 4 plots show similar trends of the 
#tooth shape with regards to tooth position. 

upperjaw<-ExtantData[ExtantData$jaw.type=="U",]
lowerjaw<-ExtantData[ExtantData$jaw.type=="L",]

#procrustes coord 7 upper
plot(upperjaw$ProcCoord7~upperjaw$tooth.position, main="tooth position/ shape Upper", 
  	xlab="position", ylab="landmark 4x")

#procrustes coord 7 lower
plot(lowerjaw$ProcCoord7~lowerjaw$tooth.position, main="tooth position/ shape Lower", 
  	xlab="position", ylab="landmark 4x")

#procrustes coord 8 upper
plot(upperjaw$ProcCoord8~upperjaw$tooth.position, main="tooth position/ shape Upper", 
  	xlab="position", ylab="landmark 4y")

#procrustes coord 8 lower
plot(lowerjaw$ProcCoord8~lowerjaw$tooth.position, main="tooth position/ shape Lower", 
  	xlab="position", ylab="landmark 4y")


```


#Clustering our Data Set; Cluster Analysis

This emphasizes both differences and similarities among individual clusters
by clustering these entities based on inter-entity resemblance. This technique
is supporting what was previously found in the CVA. It notes that the optimal
amount of clusters explain the most variation in the dataset,
which is what the CVA was able to conclude. 

This is not a necessary analysis, but just to strengthen previous results with
an alternate method.

```{r}
#find optimal number of clusters for dataset



#this function will output a log function graph showing the optimal amount
#of clusters that describe the most variation found in the dataset. 

allcoords = ExtantData[ , 7:34]
matshark <- as.matrix(allcoords)


totv <- sum(diag(var(matshark[, 1:28])))*1154
SSratio <- numeric(10)
 for (i in 2:10)
 {mod <- kmeans(matshark[, 1:28],i)
   SSratio[i]<-(totv-sum(mod$withinss))/totv}
   

plot(1:10, SSratio, xlab="number of clusters", ylab="% of explained variance")
points(1:10, c(0, SSratio[2:10]), pch=3)

#Here it is saying there are 10 clusters in order to explain the most variation. 
```



