---
title: "Shark_LDA"
output: pdf_document
---

##LDA of entire (all) matrix by species
#are some teeth better than others than distinguishing among species

```{r}

library(MASS)


upper_coord<- upper[, 6:33]
lower_coord <- lower[, 6:33]

jaw_type <-as.factor(upper[ ,4])


#make a matrix of all coords, then separate matrix to upper and lower.

matshark <- as.matrix(coords)


#overall lda

shark_lda <- lda(matshark, as.factor(sharkdat[ ,2]))
plot(shark_lda, col = 'blue')

#this does a pretty nice job of separating species out from one another
#But we can't tell what is the upper and lower jaw. Continue to do an 
#lda that distinguishes the upper and lower jaws from one another. 
#this is the opposite result of the pca; it does a better job at 
#distinguishing between species but not jaw type.

```

#analyze the lda with the upper and lower jaws separated
#LDA of Species, Upper and Lower Jaws Separated. 

```{r}

upper_coord<- upper[, 6:33]
lower_coord <- lower[, 6:33]

matshark_upper <- as.matrix(upper_coord)
matshark_lower <- as.matrix(lower_coord)

 
#Upper

upper_lda <- lda(matshark_upper, as.factor(upper[ ,2]))
plot(lda(matshark_upper, as.factor(upper[ ,2])))

#Here are the upper jaws, and the discriminant function is able to cluster #species
#distinct from one another from the upper jaw teeth. Upper jaw teeth shape #are a clasifier
#for identification to species
#we can see that the lda detected collinear variables, which is probably why it 
#couldn't distinguish between jaws and species. 


#Lower

lower_lda <- lda(matshark_lower, as.factor(lower[ ,2]))
plot(lda(matshark_lower, as.factor(lower[ ,2])))

#Here we can see that the altimus species have completely different lower #teeth
#but the lower teeth of falciformis and brevipinna have lots of overlap, 
#due to the fact they exhibit similar shape. So lower jaw teeth are not as #good
#for identifying down to the species level based on tooth shape because #they
#share more overlap. 


```

#Mahalanobis Distance Between Upper jaws species and lower jaws species

It takes into consideration the variance and correlation of the variables in measuring distances between points, i. e., differences in directions in which there is less variation within groups are given greater weight than are differences in directions in which there is more variation.

```{r}

uppermean <- upper_lda$mean
dist(predict(upper_lda, uppermean)$x)

#the mahaalanobis distance from each of the species shows they are 
#easily distinguished from one another (the larger the number, the greater 
#the greater the difference between the grous)


lowermean <- lower_lda$mean
dist(predict(lower_lda, lowermean )$x)

#here there is significant difference between Ca and Cb as expressed by the
#mahalanobis distance, along with Ca and Cf, however Cb and Cf are quite similar
#due to the small distance measurement. THis shows that there is significant overlap
#and similarities between the shape of lower jaw teeth of falciformis and #brevipinna. So there are not as good at predicting species by the shape of the
#lower jaw teeth. 

```

##Mahalanobis Distance Between Shark Species


```{r}

#now let's compute the distances between mean groups in space defined by the #dis.functions
#the compute the mahalanobis distance between groups (separating our species)

meangroup <- shark_lda$mean
dist(predict(shark_lda, meangroup)$x)

#if we compute the same for the overall LDA of the data set, we can see the same
#overlap occuring between Cf and Cb, and we can still see the clustering of Ca #separate from the other 2 species. The differences are skewed in here because
#this is combining both the shape data for the upper and lower jaws. The previous
#calculation (distance) separting out the jaw type is a better depiction 
#of analyzing the shape data. 

```


#Clustering our Data Set; Cluster Analysis

This emphasizes both differences and similarities among individual clusters
by clustering these entities based on inter-entity resemblance. 

#find optimal number of clusters for dataset

```{r}

#this function will output a log function graph showing the optimal amount
#of clusters that describe the most variation found in the dataset. 

p<-as.matrix(coords)

totv <- sum(diag(var(p[, 1:28])))*388
SSratio <- numeric(10)
 for (i in 2:10)
 {mod <- kmeans(p[, 1:28],i)
   SSratio[i]<-(totv-sum(mod$withinss))/totv}
   

plot(1:10, SSratio, xlab="number of clusters", ylab="% of explained variance")
points(1:10, c(0, SSratio[2:10]), pch=3)

#we can see here that the amount of clusters that will explain the most 
#amount of variance is about 6 (where it reaches the threshhold.)
```

#apply cluster function to dataset

will emphasize both differences and similarities among the samples by clustering
entities based on inter-group resemblance. Plot 6 clusters into the function.

```{r}
library(cluster)
library("clusterGeneration", lib.loc="~/R/win-library/3.1")
pam(p[,1:28], 6)$clustering

plot(pam(x=p[,1:28], k=6), main= "", col.p="black")

#While messy, we can see how there are 6 clusters that explain most of the 
#variance in the data set (refer to PCA loadings). These clusters show 
#those formed from the upper and lower jaws of each species. (3 species)
#Notice significant overlap which is most likely due to species similarity,
#but we can see some good separation most likely due to jaw type (which 
#does explain the most variation as seen in our RDA). So let's evaluate clustering
#due to just 2 clusters, upper and lower jaw type. So this analysis confirms
#our previous findings in the LDA analysis. Let's see if it does a better
#job when we separate just to 2 clusters, which we saw in the PCA. 


#Examine the same plot but examine Upper vs. Lower Teeth overlap
#Upper teeth are indicated with a number 2, lower teeth with a number 1. 
#Lower teeth in the first position are categorized in the upper jaw group
#because they have a indentical shape form to those in the upper jaw.
#clear separation is seen, overlap is due to similar teeth shapes of the first position teeth
#of lower jaw with upper. 

pam(x=p[,1:28], 2)$clustering
plot(pam(x=p[,1:28], k=2), main= "", col.p="black")


#confirms our previous findings that the jaws are the better descriptor 
#of variation than are species. 
#from here, using the upper jaws (since it is better to detect differences between species, let's analyze to see which landmark contributes to the most change
#between the species, so which might be the cause of shape change between these 3 #spcies. )


```


Since the upper jaw teeth are the most identifiable to species (more variable),
let's 
analyze where shape change is occuring in a single tooth relative to species.
In a previous  study, Naylor noted that the 5th position tooth in the upper 
jaw is better to distinguish among species. By looking at just one tooth in the 
upper jaw
, we are now limiting shape change now to due to species. I want to see if 
if there is visual shape change across these 3 species, and what landmarks
are associated with shape change shown by a large standard deviation. 
