---
title: "Shark Partial_9 species"
output: pdf_document
---


```{r}

#Read in the Shark data;  separating out the 3 species and jaw type for each sp.


library(vegan)
library(dummies)

#import data set (top tab in tools) data is in GIT scripts, SharkPartialData.txt

SharkPartialData <- read.delim("~/Shark-Morphometrics/Scripts/SharkPartialData.txt")
View(SharkPartialData)

```

#Subset species

```{r} 

#Subset species and jaw type.
#The 9 species below are in the genus
#Carcharhinus (requiem sharks), Rhizoprionodon, Sphyrna, and Prionance glauca.

acronotus<- SharkPartialData[SharkPartialData$species=="Ca",]
altimus<- SharkPartialData[SharkPartialData$species=="ca",]
brevipinna<- SharkPartialData[SharkPartialData$species=="Cb",]
falciformis<- SharkPartialData[SharkPartialData$species=="Cf",]
longimanus<- SharkPartialData[SharkPartialData$species=="Cg",]
leucas<- SharkPartialData[SharkPartialData$species=="Cu",]
prionace<- SharkPartialData[SharkPartialData$species=="Pg",]
lewini<- SharkPartialData[SharkPartialData$species=="Sl",]
terra<- SharkPartialData[SharkPartialData$species=="Rt",]

#subset of upper and lower jaws (includes all species)

upperjaw<-SharkPartialData[SharkPartialData$jaw.type=="U",]
lowerjaw<-SharkPartialData[SharkPartialData$jaw.type=="L",]

upper_coords <- upperjaw[, 7:34]
lower_coords <- lowerjaw[, 7:34]

#as.factor jaw and tooth from the entire dataset

jaw <- as.factor(SharkPartialData$jaw.type) #2 levels, upper and lower
tooth <- as.factor(SharkPartialData$tooth.position) #15 levels, tooth position

```


#Loop to edit data frame into matrix

```{r}

#fucntion for transforming data frame into a matrix

get_xydat = function(ct_mat) {
    if (class(ct_mat) != 'matrix') {
        ct_mat = as.matrix(ct_mat)
        print('Converting input to a matrix')
    }
    nmarkers =  ncol(ct_mat) /2 
    xy_mat = matrix(NA, nrow= nrow(ct_mat) * nmarkers, ncol=4)
    colnames(xy_mat) = c('indiv', 'landmark', 'x','y')
    startrow = 1
    for(i in 1:nrow(ct_mat)) { 
        x = ct_mat[i, seq(1, ncol(ct_mat) - 1, 2)]
        y = ct_mat[i, seq(2, ncol(ct_mat), 2)]
        endrow = i * nmarkers
        xy_mat[startrow:endrow , ] = cbind(rep(i, nmarkers), 1:nmarkers, x,y)
        startrow = endrow + 1
    }
    xy_mat = as.data.frame(xy_mat)
    return(xy_mat)
}

```

#All Species Coordinate Data

```{r}

#all coordinates for entire shark data set

allcoords = SharkPartialData[ , 7:34]
all_coords <- get_xydat(allcoords)

head(all_coords) 

```

#What is the effect of species, jaw type, and tooth position on overall variation?

```{r}

#run an RDA to look at variance partitioning

partial_rda <- rda(allcoords ~ SharkPartialData$species + as.factor(SharkPartialData$jaw.type) +
                 as.factor(SharkPartialData$tooth.position) + as.factor(SharkPartialData$replicate))


anova(partial_rda, permutations = 1000)
anova(partial_rda, by="margin")

#output shows that jaw type accounts for a significant amount of variation, followed 
#by species and then tooth position. The effect of replicate is so sublte compared to the biological
#variation being tested, that digitizing error is not considered to have any effect on tooth shape analysis.

#Further tests will only look at species, jaw type and tooth position for having any effect on tooth shape. 

RsquareAdj(partial_rda)

#variance explained by the RDA is about 65%. This is really good. 

plot(partial_rda, type='n', scaling=1)
orditorp(partial_rda, display='sp', cex=0.5, scaling=1, col='blue')
text(partial_rda, display='bp', col='red')

#VarPart for Variance Partitioning

spec_dum = dummy(SharkPartialData$species)
jawty_dum = dummy(SharkPartialData$jaw.type)
tooth_dum = dummy(SharkPartialData$tooth.position)

varpart(allcoords, spec_dum, jawty_dum, tooth_dum)
#this shows that tooth position explains more variation than species; jaw type> tooth position> species
showvarparts(3)

```

----------------------------------------
Ordination: CVA


##Cva using Package Morpho; CVA of Upper and lower jaws as a predictor

(this is what the RDA states as explaining the most variance)

```{r}

install.packages("Morpho")
library(Morpho)

allcoords = SharkPartialData[ , 7:34]

factor_shark <- as.factor(paste(SharkPartialData$species, SharkPartialData$jaw.type, sep="_"))

#RUN the CVA

cva_shark <- CVA(allcoords, groups=factor_shark, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = FALSE, p.adjust.method = "none")

cva_shark$Var
cva_shark$Dist

# get the typicality probabiliites and classifications - tagging specimens with a low probability < 0.01
#as an outlier (unassigned) 

typprob <- typprobClass(cva_shark$CVscores,groups=factor_shark)
print(typprob)

#classification accuracy is ~80%. It correctly assigned teeth to species
#80% of the time. 

```

#Plotting the CV

```{r}

install.packages("car")
require(car)

#visualize CV scores by their groups 

plot(cva_shark$CVscores, col=factor_shark, pch=as.numeric(factor_shark), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_shark$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_shark$Var[2,1],1),"%")))

text(cva_shark$CVscores, as.character(factor_shark), col=as.numeric(factor_shark), cex=.7)

#add elipses
     for(ii in 1:length(levels(factor_shark))){
        dataEllipse(cva_shark$CVscores[factor_shark==levels(factor_shark)[ii],1],
          cva_shark$CVscores[factor_shark==levels(factor_shark)[ii],2], 
          add=TRUE,levels=.80, col=c(1:7)[ii])
  }

#This does a much better job than the PC in distinguishing groups of species
#for the upper and lower jaws. It also explains 20% more variance than the first
#two pC's do. 
----
```

This looks really messy so let's divide the species up a little bit.


CV: non Car species
```{r}

#Look at the 3 species that are not Carcharhinus (Prionace, Rhizo, and Sphyrna)

allcoords = SharkPartialData[ , 7:34]
#edit out the rows that are Carcharhinus
noncar<-SharkPartialData[-c(1:1181, 1230:1550, 1629:1964), ]

#new data from of just Prionace, Rhizo, and Sphyrna

View(noncar)
noncarcoords <- noncar[ ,7:34]


factor_noncar <- as.factor(paste(noncar$species, noncar$jaw.type, sep="_"))

#RUN the CVA

cva_nocar <- CVA(noncarcoords, groups=factor_noncar, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = FALSE, p.adjust.method = "none")

cva_nocar$Var
cva_nocar$Dist


#visualize CVA

plot(cva_nocar$CVscores, col=factor_noncar, pch=as.numeric(factor_noncar), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_nocar$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_nocar$Var[2,1],1),"%")))

text(cva_nocar$CVscores, as.character(factor_noncar), col=as.numeric(factor_noncar), cex=.7)

#add elipses
     for(ii in 1:length(levels(factor_noncar))){
        dataEllipse(cva_nocar$CVscores[factor_noncar==levels(factor_noncar)[ii],1],
          cva_nocar$CVscores[factor_noncar==levels(factor_noncar)[ii],2], 
          add=TRUE,levels=.80, col=c(1:7)[ii])
  }

  
  
```

Now let's just look at just Carcharhinus species

```{r}
#data frame of just Carcharhinus species

car<-SharkPartialData[c(1:1181, 1230:1550, 1629:1964), ]

carcoords <- car[ ,7:34]

factor_car <- as.factor(paste(car$species, car$jaw.type, sep="_"))



#RUN the CVA

cva_car <- CVA(carcoords, groups=factor_car, weighting = TRUE, plot = TRUE,
             rounds = 0, cv = FALSE, p.adjust.method = "none")

cva_car$Var
cva_car$Dist


#visualize CVA

plot(cva_car$CVscores, col=factor_car, pch=as.numeric(factor_car), typ="n",asp=1,
xlab=paste("1st canonical axis", paste(round(cva_car$Var[1,1],1),"%")),
ylab=paste("2nd canonical axis", paste(round(cva_car$Var[2,1],1),"%")))

text(cva_car$CVscores, as.character(factor_car), col=as.numeric(factor_car), cex=.7)

#add elipses
     for(ii in 1:length(levels(factor_car))){
        dataEllipse(cva_car$CVscores[factor_car==levels(factor_car)[ii],1],
          cva_car$CVscores[factor_car==levels(factor_car)[ii],2], 
          add=TRUE,levels=.80, col=c(1:7)[ii])
  }


















```




```{r}
library(lattice)

#histogram showing the distribution of CA's among the species
#and jaw types. explains 53% of overall variaion

histogram(~cva_shark$CVscores[,1]|factor_shark,
  layout=c(1,length(levels(factor_shark))),
  xlab=paste("1st canonical axis", paste(round(cva_shark$Var[1,2],1),"%")))

#histogram showing same thing for CA 2; 16.4% of overall variation

histogram(~cva_shark$CVscores[,2]|factor_shark, 
      layout=c(1,length(levels(factor_shark))),
      xlab=paste("2nd canonical axis", paste(round(cva_shark$Var[2,2],1),"%")))


#plot Mahalahobis; this will show the relationships between my groups

dendroS=hclust(cva_shark$Dist$GroupdistMaha)
dendroS$labels=levels(factor_shark)
par(mar=c(4,4.5,1,1))
dendroS=as.dendrogram(dendroS)
plot(dendroS, main='',sub='', xlab="Shark Species and Jaw",
     ylab='Mahalahobis distance')

cva_shark$Dist

# Variance explained by the canonical roots:

cva_shark$Var
# or plot it:
barplot(cva_shark$Var[,2])

#CV's do a very good job of explaining overall variation in the dataset.
#much better than the PCA. 

```

-------------------------------------------------
lDA

#analyze the lda with the upper and lower jaws separated
#LDA of Species, Upper and Lower Jaws Separated. 

```{r}
library(MASS)


upperjaw<-SharkPartialData[SharkPartialData$jaw.type=="U",]
lowerjaw<-SharkPartialData[SharkPartialData$jaw.type=="L",]

upper_coords <- upperjaw[, 7:34]
lower_coords <- lowerjaw[, 7:34]


matrix_upper <- as.matrix(upper_coords)
matrix_lower <- as.matrix(lower_coords)

 
#Upper

upperjaw_lda <- lda(matrix_upper, as.factor(upperjaw[ ,2]))
plot(lda(matrix_upper, as.factor(upperjaw[ ,2])))

#Here are the upper jaws, and the discriminant function is able to cluster #species
#distinct from one another from the upper jaw teeth. Upper jaw teeth shape #are a clasifier
#for identification to species
#we can see that the lda detected collinear variables, which is probably why it 
#couldn't distinguish between jaws and species. 


#Lower

lowerjaw_lda <- lda(matrix_lower, as.factor(lowerjaw[ ,2]))
plot(lda(matrix_lower, as.factor(lowerjaw[ ,2])))




