---
title: "Shark_Euclidean"
output: pdf_document
---

##Extract Tooth Position 5 for the Upper Jaw for 3 species
#what landmarks attribute to the most variation in shape among species (euclidean distance)

```{r}

upper<-sharkdat[sharkdat$jaw.type=="U",]
fifth <- upper[upper$tooth.position=="5",]

#extract just the 5th tooth for each species 
a5<-fifth[fifth$species=="Ca",]
b5<-fifth[fifth$species=="Cb",]
f5<-fifth[fifth$species=="Cf",]


#just coordinates
a5_coords <- a5[ ,6:33]
b5_coords <- b5[ ,6:33]
f5_coords <- f5[ ,6:33]

#mean of all 5th position
m5_coords <- fifth[ ,6:33]

#xy data matrix
xydat_a5 <- get_xydat(a5_coords)
xydat_b5 <- get_xydat(b5_coords)
xydat_f5 <- get_xydat(f5_coords)

xydat_m5 <- get_xydat(m5_coords)

#altimus
x_avg_a5 <- tapply(xydat_a5$x, xydat_a5$landmark, mean)
y_avg_a5 <-tapply(xydat_a5$y, xydat_a5$landmark, mean)

#brevipinna
x_avg_b5 <- tapply(xydat_b5$x, xydat_b5$landmark, mean)
y_avg_b5 <-tapply(xydat_b5$y, xydat_b5$landmark, mean)

#falciformis
x_avg_f5 <- tapply(xydat_f5$x, xydat_f5$landmark, mean)
y_avg_f5 <-tapply(xydat_f5$y, xydat_f5$landmark, mean)


#mean
x_avg_m5 <- tapply(xydat_m5$x, xydat_m5$landmark, mean)
y_avg_m5 <- tapply(xydat_m5$y, xydat_m5$landmark, mean)
```


#Plot the 5th position for each species

```{r}

#altimus
plot(x_avg_a5, y_avg_a5, pch=19, col='red') #mean plotted in red

#brevi
points(x_avg_b5, y_avg_b5, pch=19, col='blue')

#falci
points(x_avg_f5, y_avg_f5, pch=19, col='green')

#mean shape of all 
points(x_avg_m5, y_avg_m5, pch=19, col='grey')

```

Here we can see shape changes associated with a single tooth position (5)of the 
upper jaw across 3 species and their deviation away from the mean shape. THe 
mean was calculated as the average of all the 5th position teeth from the 
3 species put together. 

##Calculate the distance between those landmarks

The distance between two landmarks is the square root of the sum of the 
squared differences between each landmark. (Euclidean distance)
This allows us to calculate a single value distance from one landmark to the 
next in each species, so then we can use these to compare to the next species.
We are going to calculate the Euclidean distance because it is a coordinate free approach for comparing biological shapes using landmark data.

```{r}

#matrix of each species, upper jaw

alti <- cbind(x_avg_a5, y_avg_a5) #red
brevi <- cbind(x_avg_b5, y_avg_b5) #blue
falci <- cbind(x_avg_f5, y_avg_f5) #grenen
mean_5 <- cbind(x_avg_m5, y_avg_m5) #grey

#calculate the interlandmark distance between each landmark, for each species
#and for the mean
#(i.e: distance between landmark 1>2>3>4>5>6...14)

dist_alti <- vegdist(alti, method="euclidean")
dist_alti <- vegdist(brevi, method="euclidean")
dist_alti <- vegdist(falci, method="euclidean")

dist_mean <- vegdist(mean_5, method="euclidean")

```

compare to see how these distances differ from the mean (which landmarks attribute
to the greatest change per/ species.) 

##Euclidean distance of species from mean (upper jaw, 5th pos.)

I want to see what landmarks are attributing to the most change in tooth shape
for each species ( from mean reference shape, to target shape:species)

```{r}

#difference of altimus from mean
apply(mean_5 - alti,1,function(x){sqrt(x[1]^2+x[2]^2)})

#difference of brevipinna from mean
apply(mean_5 - brevi,1,function(x){sqrt(x[1]^2+x[2]^2)})

#difference of falciformis from mean
apply(mean_5 - falci,1,function(x){sqrt(x[1]^2+x[2]^2)})


adist = apply(mean_5 - alti,1,function(x){sqrt(x[1]^2+x[2]^2)})
bdist = apply(mean_5 - alti,1,function(x){sqrt(x[1]^2+x[2]^2)})
bdist = apply(mean_5 - brevi,1,function(x){sqrt(x[1]^2+x[2]^2)})
fdist = apply(mean_5 - falci,1,function(x){sqrt(x[1]^2+x[2]^2)})
cbind(adist,bdist,fdist)

apply(cbind(adist,bdist,fdist),1,sd)

#so we can see that the 4th landmark (the tip of the blade) contributes to
#the most change in shape with regards to species in the upper jaw.
#we can see a considerable amount of variation due to landmark 3, 10, and 13 as well. 

```